---
title: "DAPC"
output: html_document
date: "2023-05-05"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 1. Load packages

```{r, message = FALSE, warning = FALSE}
library(DESeq2)
library(tidyverse)
library(adegenet)
```

\

## 2. Data transformation

\

##### **2.a. Import data sets**

The following data sets are provided by Dr. Rebecca L. Young.

```{r, message = FALSE}
frog_raw_counts_greaterthan5 <- read.table(file = "/stor/work/Bio321G_RY_Spring2023/MiniProjects/PoisonFrogLivers/genecounts.greaterthan5.raw.csv", 
                                           row.names = 1,
                                           header = TRUE, 
                                           sep = ",") 

frog_metadata <- read.delim(file = "/stor/work/Bio321G_RY_Spring2023/MiniProjects/PoisonFrogLivers/Metadata_EpipedobatesTranscriptomeProject2021.txt")
```

\

##### **2.b. Normalization**

Since the gene length is not provided, we will use DeSeq2 to normalize our raw counts

```{r, message = FALSE, warning = FALSE}
dds <- DESeqDataSetFromMatrix(countData=frog_raw_counts_greaterthan5, design = ~condition,
                              colData=frog_metadata)

dds <- estimateSizeFactors(dds)

normalized_counts_greater_5 <- counts(dds, normalized = TRUE)
normalized_counts_greater_5 <- as.data.frame(normalized_counts_greater_5)   # Convert to data.frame
```

\

## 3. Discriminant Analysis Principal Component (DAPC)

\

DAPC can be used to infer the number of clusters of genetically related individuals. In this multivariate statistical approach variance in the sample is partitioned into a between-group and within- group component, in an effort to maximize discrimination between groups. In DAPC, data is first transformed using a principal components analysis (PCA) and subsequently clusters are identified using discriminant analysis (DA).

In other words, the goal of DAPC is to find a linear combination of features that characterizes or separates two or more classes of objects or events. While PCA focuses on overall variation, DAPC allows us to focus specifically on finding variation between groups.


\

##### **3.a. Calculate the PCs**

```{r}
x <- normalized_counts_greater_5 %>%
  t()   # Transpose the matrix

PC_x <- prcomp(x)   # Calculate the PCs

PCs_x <- data.frame(PC_x$x) %>%
  rownames_to_column(var = "sample_id") 

head(PCs_x, 10)
```

\

##### **3.b. Data transformation**

```{r, message = FALSE, warning = FALSE}
# Remove rows containing the two anomalies using the results of the PCA
PCs_x <- PCs_x %>% 
  subset(sample_id != c("H2.6848_S16", "A3.6830_S17"))

normalized_counts_greater_5 <- normalized_counts_greater_5 %>% 
  dplyr::select(-c("H2.6848_S16", "A3.6830_S17"))

# Add another column which provides the species related to each gene
PCs_x <- PCs_x %>% 
  mutate(Species = case_when(
    sample_id == "A1.6854_S1" ~ "E_anthonyi", 
    sample_id == "B1.6855_S2"~ "E_anthonyi", 
    sample_id == "C1.6872_S3" ~ "E_anthonyi", 
    sample_id ==  "D1.6863_S4"~ "E_anthonyi", 
    sample_id == "E1.6870_S5"~ "E_anthonyi", 
    sample_id == "F1.6803_S6" ~ "E_boulengeri", 
    sample_id == "G1.6806_S7" ~ "E_boulengeri", 
    sample_id == "H1.6807_S8"~ "E_boulengeri", 
    sample_id == "B2.6813_S10" ~ "E_boulengeri", 
    sample_id ==  "C2.6821_S11"~ "E_machalilla", 
    sample_id == "D2.6822_S12"~ "E_machalilla", 
    sample_id == "E2.6826_S13" ~ "E_machalilla",
    sample_id == "F2.6845_S14" ~ "E_machalilla",
    sample_id == "G2.6847_S15_1" ~ "E_machalilla",
    sample_id == "H2.6848_S16" ~ "E_machalilla",
    TRUE ~ "E_tricolor"))   # The remaining ones are E_tricolor

# Add another column which provides the skin coloration related to each species
PCs_x <- PCs_x %>% 
  mutate(Skin_Coloration = case_when(
    Species == "E_anthonyi" ~ "Aposematic", 
    Species == "E_tricolor" ~ "Aposematic", 
    Species == "E_machalilla" ~ "Cryptic",
    Species == "E_boulengeri" ~ "Cryptic"))

# Add another column which provides the localities at which the Epipedobates were collected
PCs_x <- PCs_x %>% 
  mutate(Locality = case_when( 
    sample_id == "A1.6854_S1" ~ "Pasaje", sample_id == "B1.6855_S2" ~ "Pasaje", 
    sample_id == "C1.6872_S3" ~ "Uzchurummi", sample_id ==  "D1.6863_S4" ~ "Moromoro", 
    sample_id == "E1.6870_S5" ~ "Uzchurummi", sample_id == "F1.6803_S6" ~ "LaPerla", 
    sample_id == "G1.6806_S7" ~ "LaPerla", sample_id == "H1.6807_S8" ~ "LaPerla", 
    sample_id == "B2.6813_S10" ~ "Bilsa", sample_id ==  "C2.6821_S11"~ "5 de agosto", 
    sample_id == "D2.6822_S12" ~ "5 de agosto", sample_id == "E2.6826_S13" ~ "5 de agosto",
    sample_id == "F2.6845_S14" ~ "Jouneche", sample_id == "G2.6847_S15_1" ~ "Jouneche",
    sample_id == "H2.6848_S16" ~ "Jouneche", sample_id == "A3.6830_S17" ~ "Guanujo",
    sample_id == "C3.6835_S19" ~ "Guanujo", sample_id == "B3.6832_S18" ~ "Guanujo",
    sample_id == "D3.6843_S20" ~ "ChazoJuan"))

# Add another column which provides the sex of the Epipedobates
PCs_x <- PCs_x %>% 
  mutate(Sex = case_when( 
    sample_id == "A1.6854_S1" ~ "F", sample_id == "B1.6855_S2"~ "M", 
    sample_id == "C1.6872_S3" ~ "F", sample_id ==  "D1.6863_S4"~ "M", 
    sample_id == "E1.6870_S5"~ "F", sample_id == "F1.6803_S6" ~ "F", 
    sample_id == "G1.6806_S7" ~ "F", sample_id == "H1.6807_S8"~ "F", 
    sample_id == "B2.6813_S10" ~ "F", sample_id ==  "C2.6821_S11"~ "F", 
    sample_id == "D2.6822_S12"~ "M", sample_id == "E2.6826_S13" ~ "M",
    sample_id == "F2.6845_S14" ~ "F",sample_id == "G2.6847_S15_1" ~ "F",
    sample_id == "H2.6848_S16" ~ "F",sample_id == "A3.6830_S17" ~ "F",
    sample_id == "C3.6835_S19" ~ "F",sample_id == "B3.6832_S18" ~ "F",
    sample_id == "D3.6843_S20" ~ "M"))

# Make first column as row names 
PCs_sample <- PCs_x[,-1]
rownames(PCs_sample) <- PCs_x[,1]
```

\

##### **3.c. Calculate a-score**

To retain the optimal number of PCs, we'll run the a-score optimization (*Note*: The number of discriminant analysis `n.da` = n -1, where n is the number of different groups (e.g., species (n = 4), sex (n = 2)))

```{r, message = FALSE, warning = FALSE, fig.height = 5}
set.seed(100) # Since DAPC analysis use randomized groups to calculate a.score, we'll set seed

# Calculate optimal number of PCs for species 
dapcTemp_species <- dapc(t(normalized_counts_greater_5), PCs_sample$Species, 
                         perc.pca = 100, n.da = 3)   # n.da = 4(species) - 1 = 3
ascore_species <- optim.a.score(dapcTemp_species, smart = FALSE, n.sim = 50)   # Optimal number of PCs: 4


# Calculate optimal number of PCs for skin coloration 
dapcTemp_skin_coloration <- dapc(t(normalized_counts_greater_5), PCs_sample$Skin_Coloration, 
                                 perc.pca = 100, n.da = 1)   # n.da = 2(skin coloration) - 1 = 1
ascore_skin_coloration <- optim.a.score(dapcTemp_skin_coloration, smart = FALSE, n.sim = 50)   # Optimal number of PCs: 3


# Calculate optimal number of PCs for locality
dapcTemp_skin_locality <- dapc(t(normalized_counts_greater_5), PCs_sample$Locality, 
                                 perc.pca = 100, n.da = 18)   # n.da = 19 (localities) - 1 = 18
ascore_locality <- optim.a.score(dapcTemp_skin_locality, smart = FALSE, n.sim = 50)   # Optimal number of PCs: 3


# Calculate optimal number of PCs for sex
dapcTemp_sex <- dapc(t(normalized_counts_greater_5), PCs_sample$Sex, 
                  perc.pca = 100, n.da = 1)   # n.da = 2(sex) - 1 = 1
ascore_sex <- optim.a.score(dapcTemp_sex, smart = FALSE, n.sim = 50)   # Optimal number of PCs: 1
```

\

## 4. Visualization - DAPC

\

##### **4.a. DAPC by species**

```{r}
{ # Wrap output (e.g., plots and text) in one chunk using {}
  set.seed(100)

  dapc_species <- dapc(t(normalized_counts_greater_5), PCs_sample$Species, 
                       n.pca = 4, n.da = 3)
  scatter.dapc(dapc_species, scree.pca = TRUE, scree.da = TRUE, legend = TRUE)
  loadingplot(dapc_species$var.contr)  # var.contr: A data.frame giving the contributions of 
                                       # original variables to the principal components of DAPC
  print(paste("The proportion of conserved variance is", dapc_species$var))
}
```

We can see that the DAPC clearly separates the four species and up to 99% of the variance is conserved, performing a much better job than the PCA. If we looked at the loading plot, we can see that the two genes DN415398 and DN420333 are the main genes that drive the separation. However, the gene DN415398, which has the mapping P4HB_MAN1A1_HKDC1_HK3_HK2_HK1, are found in every category in our GO analysis, suggesting that it's a common gene. On the other hand, the gene DN420333, which has the mapping SERPINA12_SERPINA10, is not identified in our DGE or GO analysis.

\

##### **4.b. DAPC by skin coloration**

```{r}
{
  set.seed(100)

  dapc_skin_coloration <- dapc(t(normalized_counts_greater_5), PCs_sample$Skin_Coloration, 
                               n.pca = 3, n.da = 1)
  scatter.dapc(dapc_skin_coloration, scree.pca = TRUE, scree.da = TRUE, legend = TRUE)
  loadingplot(dapc_skin_coloration$var.contr) 
  print(paste("The proportion of conserved variance is", dapc_skin_coloration$var))
}
```

Again, we can see that the DAPC clearly separates the two skin coloration and up to 98% of the variance is conserved. If we looked at the loading plot, we can see that the two genes DN415398 and DN420333 are the main genes that drive the separation. However, we found the same trend where the two genes DN415398 and DN420333 are the main genes that drive the separation again.

\

##### **4.c. DAPC by locality**

```{r}
{
  set.seed(100)

  dapc_locality <- dapc(t(normalized_counts_greater_5), PCs_sample$Locality, 
                      n.pca = 3, n.da = 18)
  scatter.dapc(dapc_locality, scree.pca = TRUE, scree.da = TRUE, legend = TRUE)
  loadingplot(dapc_locality$var.contr)  
  print(paste("The proportion of conserved variance is", dapc_locality$var))
}
```

Same pattern as above

\

##### **4.d. DAPC by sex**

```{r}
{
  set.seed(100)

  dapc_sex <- dapc(t(normalized_counts_greater_5), PCs_sample$Sex, 
                   n.pca = 1, n.da = 1)
  scatter.dapc(dapc_sex, scree.pca = TRUE, scree.da = TRUE, legend = TRUE)
  loadingplot(dapc_sex$var.contr)  
  print(paste("The proportion of conserved variance is", dapc_sex$var))
}
```

This time, the separation between male and female is clear and the proportion of conserved variance remains high (although slightly lower than previous ones), the genes responsible for driving the separation are different - DN427285, DN420122, and DN404150. However, DN 427285 doesn’t have a mapping, while both DN420122 (mapping: DROSHA) and DN404150 (mapping: OMP_KIAA0226) are not identified in our DGE or GO analysis

\

##### **Overall, the genes that primarily drive the difference between the 4 groupings (species, skin coloration, locality, and sex) are either common genes, not found in our DGE or GO analysis (conflicting result), or doesn't have a mapping. Therefore, it's not possible to pinpoint any specific gene(s) that results in the separation between Cryptic and Aposematic before further investigations are carried out.**
