---
title: "PCA"
output: html_document
date: "2023-05-04"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

\

## 1. Load packages

```{r, message = FALSE, warning = FALSE}
library(tidyverse)
library(DESeq2)
library(ggrepel)
```

\

## 2. Data transformation

\

##### **2.a. Import data sets**

The following data sets are provided by Dr. Rebecca L. Young.

```{r, message = FALSE}
frog_raw_counts_greaterthan5 <- read.table(file = "/stor/work/Bio321G_RY_Spring2023/MiniProjects/PoisonFrogLivers/genecounts.greaterthan5.raw.csv", 
                                           row.names = 1,
                                           header = TRUE, 
                                           sep = ",") 

frog_metadata <- read.delim(file = "/stor/work/Bio321G_RY_Spring2023/MiniProjects/PoisonFrogLivers/Metadata_EpipedobatesTranscriptomeProject2021.txt")
```

\

##### **2.b. Normalization**

Since the gene length is not provided, we will use DeSeq2 to normalize our raw counts

```{r, message = FALSE, warning = FALSE}
dds <- DESeqDataSetFromMatrix(countData=frog_raw_counts_greaterthan5, design = ~condition,
                              colData=frog_metadata)

dds <- estimateSizeFactors(dds)

normalized_counts_greater_5 <- counts(dds, normalized = TRUE)

```

\

## 3. Principal Component Analysis (PCA)

\

PCA is a dimensionality reduction method which reduces the number of dimensions of multi-dimensional data sets, which helps us to visualize and interpret the data much better. While reducing the dimension, PCA still preserves the amount of information, allowing a comprehensive overview of the data set.

\

##### **3.a. Calculate the PCs**

```{r}
x <- normalized_counts_greater_5 %>%
  t()   # Transpose the matrix

PC_x <- prcomp(x)   # Calculate the PCs

PCs_x <- data.frame(PC_x$x) %>%
  rownames_to_column(var = "sample_id") 

head(PCs_x, 10)
```

\

##### **3.b. Scree plot**

A common method for determining the number of PCs to be retained is a graphical representation known as a scree plot. 

```{r, message = FALSE, fig.height = 6}
# Calculate the variation explained by each PCs
var_explained <- data.frame(PC = paste0("PC", 1:ncol(PC_x$x)),
                            var_explained=(PC_x$sdev)^2/sum((PC_x$sdev)^2)) 

PC1to9_Var <- var_explained[1:9,] 

ggplot(PC1to9_Var, aes(x= PC,y = var_explained * 100, group = 1)) + 
  geom_point(size=4) +
  geom_line() +
  geom_text(aes(label = round(var_explained, 4)*100, vjust = -1)) +
  labs(title = "Scree plot", y = "Percentage variation explained", x = "PC Scores") + 
  theme_classic(base_family = "Times", 
                base_size = 14)
```

\

The scree plot criterion looks for the “elbow” in the curve and selects all components just before the line flattens out. In this scree plot, the elbow point is at PC2, indicating that we only need to focus on the first two PCs. Furthermore, the first two PCs explain for almost 93% of the variation, indicating that PCA is appropriate to use in this case

\

## 4. Visualization - PCAs

\

Now, let"s plot PCA for PC1 and PC2

\

##### **4.a. Check for anomalies**

```{r}
ggplot(data = PCs_x, aes(x = PC1, y = PC2)) +
  geom_point(size = 4) +
  geom_text_repel(aes(x = PC1, y = PC2, label = sample_id)) +
  geom_hline(yintercept = 0, linetype = "dotted") +
  geom_vline(xintercept = 0, linetype = "dotted") +
  theme_linedraw(base_family = "Times", 
                 base_size = 14)
```

\

This might look daunting at first, but the purpose is to identify the anomalies, which are the two samples `H2.6848_S16` and `A3.6830_S17`.

```{r, message = FALSE, warning = FALSE}
# Remove rows containing the two anomalies
PCs_x <- PCs_x %>% 
  subset(sample_id != c("H2.6848_S16", "A3.6830_S17"))
```

\

##### **4.b. Plot by species**


```{r}
# Add another column which provides the species related to each gene
PCs_x <- PCs_x %>% 
  mutate(Species = case_when(
    sample_id == "A1.6854_S1" ~ "E_anthonyi", 
    sample_id == "B1.6855_S2"~ "E_anthonyi", 
    sample_id == "C1.6872_S3" ~ "E_anthonyi", 
    sample_id ==  "D1.6863_S4"~ "E_anthonyi", 
    sample_id == "E1.6870_S5"~ "E_anthonyi", 
    sample_id == "F1.6803_S6" ~ "E_boulengeri", 
    sample_id == "G1.6806_S7" ~ "E_boulengeri", 
    sample_id == "H1.6807_S8"~ "E_boulengeri", 
    sample_id == "B2.6813_S10" ~ "E_boulengeri", 
    sample_id ==  "C2.6821_S11"~ "E_machalilla", 
    sample_id == "D2.6822_S12"~ "E_machalilla", 
    sample_id == "E2.6826_S13" ~ "E_machalilla",
    sample_id == "F2.6845_S14" ~ "E_machalilla",
    sample_id == "G2.6847_S15_1" ~ "E_machalilla",
    sample_id == "H2.6848_S16" ~ "E_machalilla",
    TRUE ~ "E_tricolor"))   # The remaining ones are E_tricolor


# Plot
ggplot(data = PCs_x, aes(x = PC1, y = PC2, color = Species)) +
  geom_point(size = 4) +
  geom_hline(yintercept = 0, linetype = "dotted") +
  geom_vline(xintercept = 0, linetype = "dotted") +
  theme_linedraw(base_family = "Times", 
                 base_size = 14)
```

\

We can see that there"s no clear pattern/cluster.

\

##### **4.c. Plot by skin coloration**

Let"s try a different approach by cateogorizing the points by cryptic/aposematic

```{r}
# Add another column which provides the skin coloration related to each species
PCs_x <- PCs_x %>% 
  mutate(Skin_Coloration = case_when(
    Species == "E_anthonyi" ~ "Aposematic", 
    Species == "E_tricolor" ~ "Aposematic", 
    Species == "E_machalilla" ~ "Cryptic",
    Species == "E_boulengeri" ~ "Cryptic"))

# Plot
ggplot(data = PCs_x, aes(x = PC1, y = PC2, shape = Skin_Coloration, color = Species)) +
  geom_point(size = 4) +
  geom_hline(yintercept = 0, linetype = "dotted") +
  geom_vline(xintercept = 0, linetype = "dotted") +
  theme_linedraw(base_family = "Times", 
                 base_size = 14)
```

\

Again, we can see that there"s no clear pattern/cluster.

\

##### **4.d. Plot by localities**

Another potential approach could be categorizing the points by where the samples of Epipedobates were collected. The information regarding the localities is provided by Dr. Rebecca L. Young.

```{r}
# Add another column which provides the localities at which the Epipedobates were collected
PCs_x <- PCs_x %>% 
  mutate(Locality = case_when( 
    sample_id == "A1.6854_S1" ~ "Pasaje", sample_id == "B1.6855_S2" ~ "Pasaje", 
    sample_id == "C1.6872_S3" ~ "Uzchurummi", sample_id ==  "D1.6863_S4" ~ "Moromoro", 
    sample_id == "E1.6870_S5" ~ "Uzchurummi", sample_id == "F1.6803_S6" ~ "LaPerla", 
    sample_id == "G1.6806_S7" ~ "LaPerla", sample_id == "H1.6807_S8" ~ "LaPerla", 
    sample_id == "B2.6813_S10" ~ "Bilsa", sample_id ==  "C2.6821_S11"~ "5 de agosto", 
    sample_id == "D2.6822_S12" ~ "5 de agosto", sample_id == "E2.6826_S13" ~ "5 de agosto",
    sample_id == "F2.6845_S14" ~ "Jouneche", sample_id == "G2.6847_S15_1" ~ "Jouneche",
    sample_id == "H2.6848_S16" ~ "Jouneche", sample_id == "A3.6830_S17" ~ "Guanujo",
    sample_id == "C3.6835_S19" ~ "Guanujo", sample_id == "B3.6832_S18" ~ "Guanujo",
    sample_id == "D3.6843_S20" ~ "ChazoJuan"))

# Plot
ggplot(data = PCs_x, aes(x = PC1, y = PC2, shape = Skin_Coloration, color = Locality)) +
  geom_point(size = 4) +
  geom_hline(yintercept = 0, linetype = "dotted") +
  geom_vline(xintercept = 0, linetype = "dotted") +
  theme_linedraw(base_family = "Times", 
                 base_size = 14)
```

\

Again, we can see that there"s no clear pattern/cluster.

\

##### **4.e. Plot by sex**

We can also try plotting the Epipedobates by sex (i.e., males and females)

```{r}
# Add another column which provides the sex of the Epipedobates
PCs_x <- PCs_x %>% 
  mutate(Sex = case_when( 
    sample_id == "A1.6854_S1" ~ "F", sample_id == "B1.6855_S2"~ "M", 
    sample_id == "C1.6872_S3" ~ "F", sample_id ==  "D1.6863_S4"~ "M", 
    sample_id == "E1.6870_S5"~ "F", sample_id == "F1.6803_S6" ~ "F", 
    sample_id == "G1.6806_S7" ~ "F", sample_id == "H1.6807_S8"~ "F", 
    sample_id == "B2.6813_S10" ~ "F", sample_id ==  "C2.6821_S11"~ "F", 
    sample_id == "D2.6822_S12"~ "M", sample_id == "E2.6826_S13" ~ "M",
    sample_id == "F2.6845_S14" ~ "F",sample_id == "G2.6847_S15_1" ~ "F",
    sample_id == "H2.6848_S16" ~ "F",sample_id == "A3.6830_S17" ~ "F",
    sample_id == "C3.6835_S19" ~ "F",sample_id == "B3.6832_S18" ~ "F",
    sample_id == "D3.6843_S20" ~ "M"))

# Plot
ggplot(data = PCs_x, aes(x = PC1, y = PC2, shape = Skin_Coloration, color = Sex)) +
  geom_point(size = 4) +
  geom_hline(yintercept = 0, linetype = "dotted") +
  geom_vline(xintercept = 0, linetype = "dotted") +
  theme_linedraw(base_family = "Times", 
                 base_size = 14)
```

\

Here, we can see a clear cluster formed by the males. However, bare in mind that there's a disproportionate of males (5) compared to females (12) in this plot.

\

##### **Overall, besides a potential separation due to sex, the PCA displays no separation between Cryptic and Aposematic Epipedobates**